Cohesiveness in microbial community coalescence

### 2021-05-17 draft/draft_v2.tex
version 2 of draft added

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness
in microbial community coalescence}
\author[1,*]{Juan Diaz-Colunga}
\author[1,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology
and
Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[2]{Department of Microbial Biotechnology,
Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
\affil[3]{Physics of Living Systems, Department of Physics,
Massachusetts Institute of Technology, Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

@@

Together, these results suggest that the strength of top-down co-selection depends on
the environment where communities are assembled and coalescence takes place.
Communities stabilized with citrate as the primary supplied resource
display a strong degree of top-down cohesiveness,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to just be passively responding to their dominants
but rather playing an active role in community coalescence.

To investigate the determinants of top-down co-selection and the factors modulating its strength,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020}
for which the MicroCRM provides a good description.
We carried out our simulations enabling a large variation across species in terms of
their metabolic architectures.
This choice is discussed in \hyperref[box1]{Box~1}.
To reproduce our experimental protocol \textit{in silico},
we first generated a library of resources and two non-overlapping pools of species.
Each pool was used to seed a collection of 100 invasive and resident communities respectively,
that were allowed to stabilize through 20 growth-dilution cycles.
We then mixed these stable communities in pairs to carry out our coalescence
and dominant-dominant competition experiments.
A full description of our modeling framework can be found in the \methodsref{sim} section.
We found that the MicroCRM was able to capture a correspondence between
the head-to-head pairwise competition of dominants
and the outcome of community coalescence
(\figref[b]{fig2}, right panel).

[Maybe this ``outro'' sentence is not even necessary]
Our experiments had shown that this correlation varied
across communities assembled in the glutamine and citrate environments
(\figref[b]{fig2}, left and middle panels),
which led us to investigate the determinants of the strength and direction
of ecological co-selection in our model.

@@

\bigskip
\begin{center}
\includegraphics[width=8cm,keepaspectratio]{figs/figBox.png}
\end{center}
\bigskip

@@

Experimental evidence suggests that
the cohorts might have a strong effect in dictating coalescence outcomes.
At least in some instances,
head-to-head pairwise competition of dominants can be heavily influenced by the presence of
sub-dominant species
(\figref[a]{fig2}).
These findings led us to investigate the potential role of bottom-up ecological co-selection
(\figref[e]{fig1}, right panels), i.e. whether the dominants may be co-selected
for or against by their cohorts.
We asked whether our model could provide insight regarding the role of sub-dominant taxa in coalescence.
To test this, we ran a new set of simulations
this time invading resident communities with the dominants alone,
as opposed to whole communities invading together (\methodsref{sim}).
We compared the invasion success of the dominants in this new scenario with respect to our previous
simulations where they invaded accompanied by their cohorts.
The invasion success of the dominant was quantified by its relative abundance
in the final coalesced community.
In situations where positive bottom-up ecological co-selection is strong, we expect to see dominants
reaching higher invasion success with their cohorts than by themselves
(\figref[b]{fig3}, green shaded region).
On the other hand, a high degree of antagonistic bottom-up co-selection would result in dominants
invading more effectively alone than in the presence of their cohorts
(\figref[b]{fig3}, red shaded region).
Alternatively, if both forms of bottom-up co-selection are weak, we would see a similar invasion success
regardless of the presence or absence of the cohort
(\figref[b]{fig3}, gray shaded region).

\figref[b]{fig3} shows that, in the simulations,
many dominants could not invade on their own
(or could only do so at very low final relative abundances, below 0.1)
but were able to reach high frequencies when they were accompanied by their cohorts.
This is a sign of positive bottom-up co-selection being common and potentially very strong,
with negative bottom-up co-selection being rarely observed.
We then asked...

@@

\subsubsection*{Conclusion}

Ideas for the conclusions: results and previous theoretical and experimental work on different systems (and on the
general relationship between metabolic interactions and community structure) suggest that...
the strength and direction of ecological co-selection is determined by the underlying metabolic networks that
shape the structure and interactions of communities assembled in synthetic minimal environments
during coalescence.
These networks are modulated by the supplied primary resource in our conditions.
When top-down co-selection is weak, bottom-up positive co-selection can be very strong, while negative
co-selection remains rare in our experiments.
ALSO: we have only discussed metabolic interactions here. Other forms of facilitation are possible
(and probably more relevant in specific natural settings)... but cohesiveness FOR SURE arises from metabolic interactions
in our experiments, and we know the effect is ultimately metabolic because the MicroCRM captures it
(actually both forms of it, top-down and bottom-up).

Understanding the mechanisms underlying the responses of microbial communities to invasions is
an essential but poorly understood question in microbial ecology \cite{Rillig2015}.
Theory has suggested that communities may exhibit an emergent cohesiveness
\cite{Gilpin1994,Livingston2013,Tikhonov2016,Tikhonov2017},
leading to members of the same community recruiting one another during
community-community invasions.
Our results provide direct experimental evidence of ecological co-selection in a large number
of community coalescence experiments,
and highlight the critical role played by the rarer, sub-dominant species in the generation
of community cohesiveness.
Our data suggests that the strength and direction of ecological co-selection is modulated
by the underlying metabolic network that shapes the structure of
communities assembled in synthetic minimal conditions
\cite{Goldford2018,Estrela2020}.
The architecture of this network is dependent on the supplied primary carbon source.
These observations,
together with previous experimental results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

Further work will be necessary to clarify the relationship between
metabolic feedbacks, community cohesiveness, and
the strength and direction of ecological co-selection.
The experimental system that we introduced in this work can be easily expanded so that
large numbers of community coalescence experiments can be carried out in parallel.
It thus represents a promising tool to explore the properties of microbial community
coalescence in high throughput
and test quantitative theories about its role in microbiome assembly.

>>>>

% title, authors, affiliations
\title{Cohesiveness in microbial community coalescence}
\author[1,*]{Juan Diaz-Colunga}
\author[1,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology
and
Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[2]{Department of Microbial Biotechnology,
Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
\affil[3]{Physics of Living Systems, Department of Physics,
Massachusetts Institute of Technology, Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

@@

Together, these observations suggest that the strength of top-down co-selection depends on
the environment where communities are assembled and coalescence takes place.
Communities stabilized with citrate as the primary supplied resource
display a strong degree of top-down cohesiveness,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to just be passively responding to their dominants
but rather playing an active role in community coalescence.

To investigate the determinants of top-down co-selection and the factors modulating its strength,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} as implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020}
for which the MicroCRM provides a good description.
We carried out our simulations enabling a large variation across species in terms of
their metabolic architectures;
this choice is discussed in \hyperref[box1]{Box~1}.
To reproduce our experimental protocol \textit{in silico},
we first generated a library of resources and two non-overlapping pools of species.
Each pool was used to seed a collection of 100 invasive and resident communities respectively,
that were allowed to stabilize through 20 growth-dilution cycles.
We then mixed these stable communities in pairs to simulate our coalescence
and dominant-dominant competition experiments
(\methodsref{sim}).
We found that the MicroCRM was able to capture a correlation between
the head-to-head pairwise competition of dominants
and the outcome of community coalescence
(\figref[b]{fig2}, right panel),
further supporting the idea that top-down ecological co-selection can consistently
emerge from metabolic interactions across species.

\iffalse
[Maybe this ``outro'' sentence is not even necessary]
Our experiments had shown that this correlation varied
across communities assembled in the glutamine and citrate environments
(\figref[b]{fig2}, left and middle panels),
which led us to investigate the determinants of the strength and direction
of ecological co-selection in our model.

\fi
\clearpage

@@

\bigskip
\begin{center}
\includegraphics[width=8cm,keepaspectratio]{figs/figBox_v2.png}
\end{center}
\bigskip

@@

Our data indicates that the primary resource supplied to the communities can modulate
the effect of the cohorts in the dominants pairwise competition
(\figref[a]{fig2})
and the strength of top-down co-selection
(\figref[b]{fig2}, left and middle panels).
Our model suggests that this might be a result of the metabolic interactions
between community members, including the rarer taxa
(\hyperref[box1]{Box~1}).
To investigate the potential role of the cohorts in coalescence,
 i.e. whether the dominants may be co-selected for or against by them
(\figref[e]{fig1}, right panels),
we ran a new set of simulations
this time invading resident communities with the dominants alone
(\methodsref{sim}).
We compared the invasion success of the dominants in isolation with respect to our previous
simulations where they invaded accompanied by their cohorts.
The invasion success of the dominants was quantified by their relative abundance
in the final stabilized communities.
Whenever positive bottom-up ecological co-selection is strong, we expect to see dominants
reaching higher invasion success with their cohorts than by themselves
(\figref[b]{fig3}, green shaded region).
On the other hand, a high degree of antagonistic bottom-up co-selection would result in dominants
invading more effectively alone than in the presence of their cohorts
(\figref[b]{fig3}, red shaded region).
Alternatively, if both forms of bottom-up co-selection are weak, we would see a similar invasion success
regardless of the presence or absence of the cohort
(\figref[b]{fig3}, gray shaded region).

\figref[b]{fig3} shows that, in the simulations,
many dominants could not invade on their own
(or could only do so at very low final relative abundances, below 0.1)
but were able to reach high frequencies when they were accompanied by their cohorts.
This indicates that positive bottom-up co-selection is frequent and potentially very strong,
while negative bottom-up co-selection is far more uncommon.
We then asked whether the ability of the pairwise competition of dominants
to predict coalescence outcomes was dependent on the strength of bottom-up co-selection.
We divided our simulations into two subsets:
the first one was comprised of the instances where positive bottom-up co-selection was strong
(i.e. dots in the green shaded region of \figref[b]{fig3}),
the second set included all other cases
(dots near the diagonal of \figref[b]{fig3}).
We went back to our original simulations and plotted the frequency of the invasive dominant
in pairwise competition with the resident dominant versus the relative similarity between
the invasive and coalesced communities,
i.e. the same plot as in \figref[b]{fig2}, for each subset.
We found that when bottom-up positive co-selection is strong,
the pairwise competition of dominants is not predictive of coalescence outcomes
(\figref[c]{fig3}, left panel) and vice-versa (\figref[c]{fig3}, right panel).

We then asked whether this trend was also observed \textit{in vitro}.
We went back to our synthetic communities
and carried out a new round of experiments where we invaded the resident communities
with the invasive dominants alone
(\methodsref{competitions}).
After stabilization, we quantified species abundance through 16S Illumina sequencing
(\methodsref{sequencing}).
Again, we observed that bottom-up co-selection is far more common in its
positive than in its negative form
(\figref[d]{fig3}).
Interestingly, bottom-up recruitment appears to be more frequent in the glutamine
environments than in the citrate ones,
consistent with our hypothesis that metabolic interactions among species
are key in determining the strength and direction of ecological co-selection.
We then repeated our analysis in \figref[c]{fig3},
this time splitting our data according to the observed strength of bottom-up
co-selection instead of the primary carbon source
as we had done in \figref[b]{fig2}.
Our findings were in line with the model prediction:
pairwise competition between dominants is only predictive of coalescence outcomes
if bottom-up co-selection is weak.

@@

\subsubsection*{Conclusions}

\iffalse
Ideas for the conclusions: results and previous theoretical and experimental work on different systems (and on the
general relationship between metabolic interactions and community structure) suggest that...
the strength and direction of ecological co-selection is determined by the underlying metabolic networks that
shape the structure and interactions of communities assembled in synthetic minimal environments
during coalescence.
These networks are modulated by the supplied primary resource in our conditions.
When top-down co-selection is weak, bottom-up positive co-selection can be very strong, while negative
co-selection remains rare in our experiments.
ALSO: we have only discussed metabolic interactions here. Other forms of facilitation are possible
(and probably more relevant in specific natural settings)... but cohesiveness FOR SURE arises from metabolic interactions
in our experiments, and we know the effect is ultimately metabolic because the MicroCRM captures it
(actually both forms of it, top-down and bottom-up).
\fi

Understanding the mechanisms underlying the responses of microbial communities to invasions is
an essential but poorly understood question in microbial ecology \cite{Rillig2015}.
Theory has suggested that communities may exhibit an emergent cohesiveness
\cite{Gilpin1994,Livingston2013,Tikhonov2016,Tikhonov2017},
leading to members of the same community recruiting one another during
community-community invasions.
Our results provide direct experimental evidence of ecological co-selection in a large number
of community coalescence experiments,
and highlight the critical role played by the rarer, sub-dominant species in the generation
of community cohesiveness.

Our data suggests that the strength and direction of ecological co-selection is modulated
by the underlying metabolic networks that shape the structure of
communities assembled in synthetic minimal conditions
\cite{Goldford2018,Estrela2020}.
This network is in turn regulated by the supplied primary carbon source.
This idea is supported by the observation that a Microbial Consumer-Resource Model
captures the trends observed experimentally when we enable a large variation
in the metabolic fluxes across species.
The model also predicts a trade-off between the strength of bottom-up co-selection
and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
which we have confirmed experimentally.
These observations,
together with previous results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

Additional work will be necessary to further clarify the relationship between
metabolic feedbacks, community cohesiveness and ecological co-selection.
The experimental system that we introduced in this work can be easily expanded so that
large numbers of community coalescence experiments can be carried out in parallel.
It thus represents a promising tool to explore the properties of microbial community
coalescence in high throughput
and test quantitative theories about its role in microbiome assembly.
On the other hand, coalescence of communities under different settings
(e.g. in spatially structured environments)
might be governed by additional factors.
Understanding them and quantifying their relative contributions in natural communities
remains an open question in ecology.
